Regression Cube Analysis of Population Study Data

### 2015-06-21 draft.tex
changed title

\title{Regression Cube Analysis of Cohort Study Data}

@@

%other entries to be set up for journal
\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Regression Cube Analysis of Cohort Study Data}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical condition as well as medical image data.
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set w.r.t. a target feature, no suitable tool is available for epidemiological researchers.

@@

We described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the feature listing step \cite{Klemm2014VIS}.
%%
The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypothesis generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.
%%
In this work, we focus on a similar approach to derive insight and even new hypotheses through the data rather than only using it for a confirmatory analysis.
%%
\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
There are many data acquisition modalities available to epidemiologists, which are suitable for different analysis types.
%%
We focus on cohort studies, which impose the largest data sets and yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either regarding a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed w.r.t. different diseases.
%%
The feature space comprises information about lifestyle, somatometric features, medical parameters, genetic data as well as medical images.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow-up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
Medical status features or lifestyle factors are primarily of \emph{dichotomous} (binary) type.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the feature types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space.
%%
It also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data are also acquired and analyzed in modern cohort studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of cohort study data types and characteristics can be found in the works of Preim et al.~\cite{Preim2014} and Toennies et al.~\cite{Toennies2015}.
%%
\\\\
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%%
The analysis model has to be adjusted by normalizing all included features w.r.t. the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely for older subjects.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.
%%
\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per cohort subject.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study.
%%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data are too often analyzed solely using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses instead, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
\\\\
%The work of M\"{u}hlbacher et al.~\cite{Muehlbacher}, Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
\paragraph{Parameter Space Analysis using Regression Models.}
%%
\magenta{
Sedlmair et al.~\cite{Sedlmair} present a conceptual framework of parameter space analysis.
%%
%It includes a abstract application domain independent data flow model, navigation strategies for exploring parameter spaces as well a characterization of analysis tasks.
%%
The taxonomy allows to sort approaches to find works with similar problems to derive new design ideas and evaluate ideas.
%%
Based on their input parameter taxonomy, we are visualizing \emph{model parameters} based on \emph{environmental parameters} in a \emph{global-to-local} navigation strategy.
%%
We solve a \emph{fitting} task by aiming to find models well suited for describing the input data.
}

@@

%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al.~\cite{Niemann2014} investigate risk factors of hepatic steatosis using decision trees with interactive data mining tools.
%%
They extract classification rules that serve as basis for our proof-of-concept tests.
%%
Niemann et al.~\cite{Niemann2015} improved the classification performance by generating features (called \emph{evolution features}) that describe latent temporal information across the study waves.
%%
We try to reproduce their results and investigate findings presented by Niemann et al.~\cite{Niemann2014} further.
%%
\paragraph{Prior Work.}
In our prior work, we analyzed the healthy aging process of the lumbar spine.
%%
%We utilized semi-automatic algorithms do detect the lumbar spine shape \cite{Klemm2013VMV, Rak2013}.
%%
%The reliable part of the presented models are represented by the shape of the lumbar spine canal, which was extracted and described using line representation.
%%
We defined an Interactive Visual Analysis workflow for image-centric cohort study data \cite{Klemm2014VIS}. 
%%
We extended the feature selection step (recall Sec. \ref{sec:Background}) with an iterative analysis loop incorporating group selection and visualization using expert input as well as clustering methods.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image and non-image information.
%%
The mosaic plot \cite{Klemm2014VIS} highlights the pairwise correlation between features without a target disease by depicting the \emph{Cram\'{e}r's V} contingency values.
%%
Its great popularity among our domain experts was the inspiration for this work.
%%

@@

\section{Regression Cube Analysis of Cohort Study Data}

@@

%%
\textbf{Attention Steering is Crucial.}
%%
Important events have to be highlighted in overview visualizations to direct the user's attention to interesting parts of the data.
%%
Poor guidance potentially leads to overlooked relationships.
%%
We found supporting mini-map visualizations, such as the 3D cube, most useful for this purpose, e.g., for highlighting differences rather than displaying absolute values (Fig.~\ref{fig:Application}).
%%
\section{Summary and Outlook}
%%
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to the most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Modeling expert knowledge using regression formulas allows for a \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for a detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.

>>>>

\title{Regression Cube Analysis of Population Study Data}

@@

%other entries to be set up for journal
\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Regression Cube Analysis of Population Study Data}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{%
%\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subject's lifestyle, medical condition as well as medical image data.
%%
Statistical regression analysis is used to evaluate these features and to identify feature combinations indicating a disease (the \emph{target feature}).
%%
Although there is a strong demand for overview visualizations of a whole data set w.r.t. a target feature, no suitable tool is available for epidemiological researchers.

@@

We described an Interactive Visual Analysis approach for image-centric cohort study data, which connects to the feature listing step \cite{Klemm2014VIS}.
%%
The methods aim to derive hypotheses through analysis of the data and observations about previously unknown feature correlations.
%%
Employing \emph{hypothesis generation} requires overview visualizations of feature correlations, which are not supported by standard statistical processors.
%%
In this work, we focus on a similar approach to derive insight and even new hypotheses through the data rather than only using it for a confirmatory analysis.
%%
\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
There are many data acquisition modalities available to epidemiologists, which are suitable for different analysis types.
%%
We focus on population studies, which impose the largest data sets and yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either regarding a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed w.r.t. different diseases.
%%
The feature space comprises information about lifestyle, somatometric features, medical parameters, genetic data as well as medical images.
%%
These features are derived through different modalities, such as questionnaires, medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow-up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.
%%
Medical status features or lifestyle factors are primarily of \emph{dichotomous} (binary) type.
%%
Continuous data are often discretized (e.g. 10 year steps for age) to equalize the feature types and to simplify the method selection.
%%
However, this should be avoided, since it reduces the information space.
%%
It also introduces a new information bias, as assumptions are modeled through the discretization.
%%

Medical image data are also acquired and analyzed in modern population studies.
%%
We incorporate image-derived features, but do not focus on analyzing image data.
%%
More discussions of population study data types and characteristics can be found in the works of Preim et al.~\cite{Preim2014} and Toennies et al.~\cite{Toennies2015}.
%%
\\\\
Features influencing the exposure as well as the outcome of an analysis are called \emph{confounders} and have to be specially treated.
%%
The analysis model has to be adjusted by normalizing all included features w.r.t. the confounder.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely for older subjects.
%%
It also influences the general body condition and thereby almost all features acquired through population studies.
%%
Another important confounder is \emph{gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

@@

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models, as opposed to linear models, do not allow for extracting an $R^2$ quality of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} behaves different than $R^2$ values extracted from the linear regression model.
%%
Comparisons have to be handled with care.
%%
\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per participant.
%%
A second cohort, \texttt{SHIP-TREND} was started in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the latest two acquisitions (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study.
%%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
In 1977, Tukey already stated that data are too often analyzed solely using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses instead, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
\\\\
%The work of M\"{u}hlbacher et al.~\cite{Muehlbacher}, Piringer et al.~\cite{Piringer}, Chan et al.~\cite{Chan} and Zhang et al.~\cite{Zhang2014} are closest related to ours regarding different aspects.
\paragraph{Parameter Space Analysis using Regression Models.}
%%
\magenta{
Sedlmair et al.~\cite{Sedlmair} present a conceptual framework of parameter space analysis.
%%
%It includes a abstract application domain independent data flow model, navigation strategies for exploring parameter spaces as well a characterization of analysis tasks.
%%
The taxonomy allows to sort approaches to find works with similar problems to derive new design ideas and evaluate ideas.
%%
Based on their input parameter taxonomy, we are visualizing \emph{model parameters} based on \emph{environmental parameters} in a \emph{global-to-local} navigation strategy.
%%
We solve a \emph{fitting} task by aiming to find models well suited for describing the input data.
}

@@

%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al.~\cite{Niemann2014} investigate risk factors of hepatic steatosis using decision trees with interactive data mining tools.
%%
They extract classification rules that serve as basis for our proof-of-concept tests.
%%
Niemann et al.~\cite{Niemann2015} improved the classification performance by generating features (called \emph{evolution features}) that describe latent temporal information across the study waves.
%%
We try to reproduce their results and investigate findings presented by Niemann et al.~\cite{Niemann2014} further.
%%
\paragraph{Prior Work.}
In our prior work, we analyzed the healthy aging process of the lumbar spine.
%%
%We utilized semi-automatic algorithms do detect the lumbar spine shape \cite{Klemm2013VMV, Rak2013}.
%%
%The reliable part of the presented models are represented by the shape of the lumbar spine canal, which was extracted and described using line representation.
%%
We defined an Interactive Visual Analysis workflow for image-centric population study data \cite{Klemm2014VIS}.
%%
We extended the feature selection step (recall Sec. \ref{sec:Background}) with an iterative analysis loop incorporating group selection and visualization using expert input as well as clustering methods.
%%
Information visualizations of population study features were augmented with extracted image data, yielding linked views with both image and non-image information.
%%
The mosaic plot \cite{Klemm2014VIS} highlights the pairwise correlation between features without a target disease by depicting the \emph{Cram\'{e}r's V} contingency values.
%%
Its great popularity among our domain experts was the inspiration for this work.
%%

@@

\section{Regression Cube Analysis of Population Study Data}

@@

%%
\textbf{Attention Steering is Crucial.}
%%
Important events have to be highlighted in overview visualizations to direct the user's attention to interesting parts of the data.
%%
Poor guidance potentially leads to overlooked relationships.
%%
We found supporting mini-map visualizations, such as the 3D cube, most useful for this purpose, e.g., for highlighting differences rather than displaying absolute values (Fig.~\ref{fig:Application}).
%%
\section{Summary and Outlook}
%%
We presented a technique for knowledge discovery in population study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to the most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Modeling expert knowledge using regression formulas allows for a \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for a detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.
